Online Supplement: \\ How important are functional and developmental constraints on phenotypic evolution? An empirical test with the stomatal anatomy of flowering plants, \\ \textit{The~American~Naturalist}

### 2022-11-01 ms/ms.Rmd
Moved online supplement to separate document for publication

---
output:
  bookdown::pdf_document2:
    includes:
      before_body: ../template/amnat_doc_prefix.tex
      in_header: ../template/amnat_preamble.tex
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: no
    toc: no
  bookdown::html_document2:
    number_sections: no
    theme: readable
    toc: yes
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::word_document2: null
fontsize: 12pt
# linestretch: 1.0
linestretch: 1.7
link-citations: yes
bst: ../template/amnat.bst
bibliography: stomata-independence-1.bib
---

@@

<!-- % For Am Nat, moved to "template/amnat_preamble.txt" -->
\begin{flushleft}
{\Large
\textbf\newline{How important are functional and developmental constraints on phenotypic evolution? An empirical test with the stomatal anatomy of flowering plants}
}
\newline
\\
Christopher D. Muir\textsuperscript{1,*},
Miquel \`{A}ngel Conesa\textsuperscript{2},
Jeroni Galm\'{e}s\textsuperscript{2},
Varsha S. Pathare\textsuperscript{3},
Patricia Rivera\textsuperscript{4},
Rosana López Rodríguez\textsuperscript{5},
Teresa Terrazas\textsuperscript{4},
Dongliang Xiong\textsuperscript{6}
\\
\bigskip
\bf{1} School of Life Sciences, University of Hawaii at M\=anoa, Honolulu, HI 96822, USA \\
\bf{2} Research Group on Plant Biology under Mediterranean Conditions, Departament de Biologia, Universitat de les Illes Balears, Ctra. Valldemossa km 7.5, E-07122, Palma, Spain \\
\bf{3} School of Biological Sciences, Washington State University, Pullman, WA 99164-4236, USA \\
\bf{4} Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Apartado Postal 70‑367, 04510 Mexico City, Mexico \\
\bf{5} Departamento de Sistemas y Recursos Naturales, Universidad Politécnica de Madrid, 28040 Madrid, Spain \\
\bf{6} National Key Laboratory of Crop Genetic Improvement, MOA Key Laboratory of Crop Ecophysiology and Farming System in the Middle Reaches of the Yangtze River, College of Plant Science and Technology, Huazhong Agricultural University, Wuhan, Hubei 430070, China\\
\bigskip
* cdmuir@hawaii.edu

@@

<!-- \textit{Manuscript elements}: Figure 1, figure 2, figure 3, figure 4, figure 5, figure 6, table 1, online appendix A (including Notes A1, notes A2, notes A3, figure A1, figure A2, figure A3, table A1, table A2, and table A3). -->

@@

\linenumbers{}
\modulolinenumbers[3]

<!-- \newpage{} -->

@@

Stomatal anatomy on the leaves of flowering plants provide an exceptional opportunity to test for functional and developmental constraints because 1) there are 1000s of species to compare and 2) the main packing constraints and developmental steps are analytically tractable. This means it is possible to derive quantitative predictions and test their generality using large comparative data sets representing millions of years of evolutionary history. For this purpose, a heretofore unappreciated fact about stomata is that many leaves have stomata on both lower and upper surfaces. The packing and developmental constraints are the same for stomata on each surface, but the selection may differ. Therefore, if packing and developmental constraints dominate, stomatal anatomy on each surface should diverge in concert. Failure to do so implies that independent evolution is possible and that selection against generally maladaptive phenotypes explains at least some of the covariance between traits. Phylogenetic comparisons of stomatal anatomy provide a statistically powerful, general, and elegant way to distinguish different phenotypic constraints that would be impossible in many other traits with as much ecological significance. The next sections provide background information on stomatal anatomy, how it varies, and why functional or developmental constraints might be important.

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/concepts.pdf}
\caption{Two salient features of stomatal anatomy in flowering plants are the (a) inverse relationship between stomatal size and density and (b) the bimodal distribution of stomatal ratio. At broad phylogenetic scales, species with smaller stomata on their leaves ($x$-axis, log-scale) tend to have greater stomatal density  ($x$-axis, log-scale), but there is a lot of variation about the overall trend indicated by the grey ellipse. Hypostomatous leaves (stomatal ratio = 0) are more common than amphistomatoues leaves, but within amphistomatous leaves, the density of stomata on each surface tends to be similar (stomatal ratio $\approx$ 0.5), which we refer to as bimodal stomatal ratio.}
\label{fig:concepts}
\end{figure}

@@

Given the significance of stomata for plant function and global vegetation modeling [@berry_stomata:_2010], we would like to understand what factors constrain their anatomical variation. Here we focus on interspecific variation in mean trait values rather than intraspecific variation. Packing constraints and developmental integration could explain inverse size-density scaling. The density and size of a stomata on a planar leaf surface can be viewed as a packing problem where the total area allocated to stomata cannot exceed the total leaf area. This is a functional, or geometric, constraint because certain combinations of large size and high density are not physically possible. This functional constraint cannot explain why combinations of low density and small size are rare, but may explain why stomatal size must decrease when density increases as the leaf runs out of space. The packing limit of functional stomata is less than the entire leaf area, but the exact value is unclear. The realized upper limit is close to 1/3 or 1/2 [@de_boer_optimal_2016; @sack_developmental_2016; @liu_scaling_2021] for the species' mean, not an individual leaf.

@@

The second hypothesis is the cell size mediates developmental integration between stomatal size and density. If developmental integration is the primary reason for inverse size-density scaling, then amphistomatous leaves will exhibit identical size-density scaling on each surface. If the stomatal size and density scale differently on each surface, this implies that they can evolve independently and that selection against generally maladaptive trait combinations shapes their covariance. Furthermore, we predicted that divergence in genome size, which is strongly associated with meristematic cell volume [@simova_geometrical_2012], would covary with stomatal size and density similarly on each surface.

@@

<!--- not using I think
In species where stomata occupy a large fraction of the epidermis, divergence will be limited by the fact that any increases in stomatal size and/or density may be deleterious because they increase stomatal area beyond an upper bound. Since less area of the adaxial (upper) surface is typically allocated to stomata, there are more ways to evolve without exceeding an upper bound in stomatal area allocation, which should increase the rate of divergence in adaxial stomatal traits. We tested this prediction with the same data set.
-->

@@

Unless otherwise mentioned, we performed all data wrangling and statistical analyses in *R* version `r as.character(getRversion())` [@r_core_team_r:_2022]. Source code is publicly available on GitHub (https://github.com/cdmuir/stomata-independence) and will be archived on Zenodo upon publication.

@@

Developmental integration could shape patterns of phenotypic macroevolution, but a major hindrance to progress is that verbal models do not make precise, quantitative predictions that distinguish it from alternatives. An advantage of testing developmental integration in stomata is that their development is well studied [@bergmann_stomatal_2007; @dow_patterning_2014; @sack_developmental_2016]. We can leverage that knowledge to build a developmental function and derive equations for phenotypic (co)variance caused by developmental integration. If observed patterns of evolution are inconsistent with developmental integration, theory may also help identify which parameters lead to developmental disintegration. We combined and extended stomatal development models to predict how stomatal density and length would diverge if stomatal development were constrained and how those predictions would change if stomatal development were unconstrained. We summarize our methods verbally here and direct readers interested in the mathematical details to Notes A1. A graphical summary is provided in Fig. \@ref(fig:developmental-integration). We imposed constraint by assuming the stomatal developmental function is constrained. The developmental function maps cell size prior to differentiation onto stomatal size and density using two parameters. The first parameter describes how cell volume is apportioned between epidermal cells and guard cell meristemoids during asymmetric cell division [@dow_patterning_2014]. The second parameter is stomatal index [@salisbury_i_1928; @sack_developmental_2016], which is determined by amplifying and spacing divisions after asymmetric cell division [@dow_patterning_2014]. When these parameters are fixed, divergence in stomatal size and density is determined by divergence in meristematic cell volume and expansion prior to asymmetric division. We relaxed this constraint by treating parameters of the developmental function as random variables that can diverge between species. We used random variable algebra to derive predicted (co)variance in divergence between stomatal length and density [see @lynch_genetics_1998 for key random variable algebra theorems].

@@

We searched the literature for studies that measured stomatal density and stomatal size, either guard cell length or stomatal pore length, for both abaxial and adaxial leaf surfaces. In other words, we did not include studies unless they reported separate density and size values for each surface. We did not record leaf angle because it is typically not reported, but we presume that for the vast majority of taxa that the abaxial is the lower surface and the adaxial is the upper surface. This is reversed in resupinate leaves, but to the best of our knowledge, our synthesis did not include resupinate leaves. None of the species with resupinate leaves listed by @chitwood_conflict_2012 are in our data set. We refer to guard cell length as stomatal length and converted stomatal pore length to stomatal length assuming guard cell length is twice pore length [@sack_developmental_2016]. Table \@ref(tab:traits) lists focal traits and symbols.

tribble(
  ~Symbol, ~`Variable string`, ~Units,
  "$D_\\mathrm{ab}$", "$\\mathtt{abaxial\\_stomatal\\_density\\_mm2}$", "$\\text{pores mm}^{-2}$",
  "$D_\\mathrm{ad}$", "$\\mathtt{adaxial\\_stomatal\\_density\\_mm2}$", "$\\text{pores mm}^{-2}$",
  "$L_\\mathrm{ab}$", "$\\mathtt{abaxial\\_stomatal\\_length\\_um}$", "$\\mu$m",
  "$L_\\mathrm{ad}$", "$\\mathtt{adaxial\\_stomatal\\_length\\_um}$", "$\\mu$m"
) |>
  knitr::kable(format = "markdown", caption = "Stomatal anatomical traits with mathemtical symbol, variable string used in source code, and scientific units.", align = "ccc")

@@

We generated an ultrametric, bifurcating phylogeny of `r n_tax3` taxa by resolving and removing ambiguous taxonomic names, placing taxa on the GBOTB.extended mega-tree of seed plants [@smith_constructing_2018; @zanne_three_2014], and resolving polytomies using published sequence data. Divergence times in this phylogeny are based on extensive fossil calibration [see @smith_constructing_2018; @magallon_metacalibrated_2015 for further detail]. The complete methodology is described in the online supplement (Notes A2).

@@

<!-- Explanation for using pairs rather than whole tree. Removed based on feeedback from Reviewer 2
We only used pairs of terminal taxa rather than the entire tree for two reasons. First, even in whole-tree methods, approximately half of the data comes from divergence at the tips, so we do not lose much statistical power. Second, using a taxon-pair method obviates making strong assumptions about multivariate trait evolution process homogeneity throughout flowering plant evolution. Homogeneity is biologically unlikely and it is difficult to fit more complex models reliably with only data from extant taxa.
-->

@@

All hypotheses make predictions about trait (co)variance matrices or parameters derived from them (see Notes A1 and subsections below). Within and among species covariation is a hallmark of developmental integration [@armbruster_multilevel_1988], but other evolutionary processes also lead to covariance. Distinguishing between them requires deriving predictions and testing whether observed covariance is consistent with one hypothesis or another. We estimated the $4 \times 4$ covariance matrix of phylogenetically independent contrasts between log-transformed values of $\Delta \text{log}(D_\mathrm{ab})$, $\Delta \text{log}(D_\mathrm{ad})$, $\Delta \text{log}(L_\mathrm{ab})$, and $\Delta \text{log}(L_\mathrm{ad})$ using a distributional multiresponse robust Bayesian approach. See Table \ref{tab:traits} for variable definitions. We denote variances as $\text{Var}[\Delta \text{log(trait)}]$ and covariances as $\text{Cov}[\Delta \text{log(trait}_1),\Delta \text{log(trait}_2)]$. We used a multivariate $t$-distribution rather than a Normal distribution because estimates using the former are more robust to exceptional trait values [@lange_robust_1989]. We also estimated whether the variance in trait divergence increases with time. Under many trait evolution models (e.g. Brownian motion), interspecific variance increases through time. To account for this, we included time since taxon-pair divergence as an explanatory variable affecting the trait covariance matrix.

@@

We asked how divergence in stomatal length and density would covary if the developmental function were constrained and compared it to their divergence when the developmental function can evolve. When the developmental function is constrained this means that allocation to guard cell meristemoids during asymmetric division and stomatal index are fixed (see Notes A1 for mathematical description). Under these assumptions, divergence in stomatal length and density is mediated entirely by divergence in meristematic cell volume and expansion prior to differentiation. Developmental integration is strong because divergence in density is perfectly negatively correlated with divergence in size. In contrast, stomatal length and density can diverge independently when the developmental function is not fixed. Divergence in asymmetric cell division affects stomatal size independently of density; divergence in stomatal index affects stomatal density independently of size. Divergence in the developmental function causes developmental disintegration because stomatal density and size can diverge independently. Developmental integration is minimal when asymmetric cell division and/or stomatal index diverge more than meristematic cell volume and expansion. The three main conclusions are that 1) developmental constraint leads to developmental integration; 2) different (co)variance in divergence of stomatal length and density on each surface implies the developmental function is not fixed; and 3) divergence in different components of the developmental function affect stomatal length and density differently. See Notes A1 and Table \@ref(tab:predictions) for more a complete derivation and detailed predictions.

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h3.pdf}
\caption{Evolutionary divergence slows down as epidermal space fills up. This pattern is consistent with functional constraints (packing limits) constraining evolution, but only when stomata occupy a large fraction of epidermal area. The shaded area in each facet indicates the estimate of where 95\% of the `r n_pair` phylogenetically independent contrasts fall as a function of the fraction of epidermal area allocated to stomata per surface. Each point is the absolute value of  $\Delta~\textrm{log(trait)}$ for stomatal density (left facets) or length (right facets) on the adaxial (upper facets) and abaxial (lower facets) surface. The fraction of epidermal area allocated to stomata is the average value per surface between the two taxa in each contrast. Divergence in anatomical traits is more variable when stomata occupy a smaller area, especially for adaxial stomatal density (upper left facet). See Table \ref{tab:h3output} for all parameter estimates and confidence intervals.}
\label{fig:h3}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h1-raw.pdf}
\caption{Inverse size-density scaling differs between the surfaces of amphistomatous leaf, indicating different phenotypic constraints. The panels show the relationship between stomatal length ($x$-axis) and stomatal density ($y$-axis) on a log-log scale for values measured on the abaxial leaf surface (left) and the adaxial leaf surface (right) across `r n_tax3` taxa.}
\label{fig:h1_raw}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h1.pdf}
  \caption{(Caption next page.)}
  \label{fig:h1}
\end{figure}
\addtocounter{figure}{-1}

\begin{figure} [t!]
\caption{(Previous page.) Greater overall phenotypic constraint on abaxial stomatal anatomy is evinced by more variable evolutionary divergence in adaxial stomatal density, but covariance between density and length is similar on both surfaces. (a) Data from `r n_pair` phylogenetically independent contrasts of change in log(stomatal length) ($x$-axis) and log(stomatal density) ($y$-axis) for abaxial (left panel) and adaxial (right panel) leaf surfaces. Each contrast is shown by black points and every contrast appears on both panels. Grey ellipses are the model-estimated 95\% covariance ellipses. The negative covariance is similar for both surfaces but the breadth in the $y$-direction is larger for adaxial traits, indicating greater evolutionary divergence in log(stomatal density). (b) Parameter estimates (points), 66\% (thick lines), and 95\% HPD intervals for estimates of trait (co)variance. Grey points and lines represent ab- and adaxial values; black points and lines represent the estimated difference in (co)variance between surfaces. Only the variance for stomatal density (middle panel) is significantly greater for the adaxial surface (95\% HPD interval does not overlap the dashed line at 0). Reported parameter estimates are conditioned on zero time since divergence between taxa (see \protect\hyperlink{results}{Results}).}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h2-raw.pdf}
\caption{Stomatal density on each surface is less constrained (lower correlation) than stomatal length (higher correlation). Relationship between stomatal density and length on each leaf surface in a synthesis of amphistomatous leaf traits across `r n_tax3` taxa. The panels show the relationship between the abaxial trait value ($x$-axis) and the adaxial trait value ($y$-axis) on a log-log scale for stomatal density (left) and stomatal length (right). The dashed line in across the middle is the 1:1 line for reference.}
\label{fig:h2_raw}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=\textwidth]{../figures/h2.pdf}
\caption{Developmental integration in stomatal length is much stronger than stomatal density between the surfaces of amphistomatous leaves (a) Data from `r n_pair` phylogenetically independent contrasts of change in the abaxial trait value ($x$-axis) against change in the adaxial trait value ($y$-axis) for log(stomatal density) (left panel) and log(stomatal length) (right panel). Each contrast is shown by black points and every contrast appears on both panels. Dashed grey lines are 1:1 lines for reference. Solid grey lines and ribbon the fitted SMA slope and 95\% HPD interval. (b) The SMA slope (left panel) is significantly less than 1 (isometry, top dashed line) for density but very close to isometric for length. The coefficient of determination ($r^2$, right panel) is also much greater for length than density. The points are parameter estimates with 66\% (thick lines) and 95\% HPD intervals. Reported parameter estimates are conditioned on zero time since divergence between taxa (see \protect\hyperlink{results}{Results}).}
\label{fig:h2}
\end{figure}

@@

Functional and developmental constraints may slow adaptation by preventing traits from evolving independently towards a multivariate phenotypic optimum. An alternative view is that phenotypic evolution is limited primarily because most phenotypes are maladaptive and selected against, sometimes referred to as selective constraint. In this study, we took advantage of the fact that amphistomatous leaves produce stomata on both abaxial (usually lower) and adaxial (usually upper) surfaces. Packing limits, a functional constraint, and developmental integration should result in similar (co)variance in divergence of stomatal traits on each surface (Notes A1), whereas differing selective pressures for each surface would lead to different patterns of divergence. Although some patterns of trait (co)variance are compatible with packing limits or developmental integration (see below), independent evolution of stomatal density on each surface across flowering plants is notably inconsistent with these constraints. We therefore conclude that selection against generally maladaptive trait combinations plays a major role in the evolution of stomata and possibly other ecologically important traits. It should not be surprising that stomatal traits respond to natural selection, but it is nevertheless plausible that some stomatal trait combinations that would have high fitness are precluded by nonadaptive constraints. Our results challenge these nonadaptive explanations, especially for stomatal density.

@@

If meristematic cell volume and expansion integrate stomatal size and density [@brodribb_unified_2013], then we predicted inverse size-density scaling would evolve with the same (co)variance for both ab- and adaxial leaf surfaces (Notes A1). Contrary to this prediction, there are many combinations of stomatal density ($D_i$) and length ($L_i$) found on adaxial leaf surfaces that are absent from abaxial leaf surfaces (Fig. \ref{fig:h1_raw}). Abaxial traits are subscripted with $_{\textrm{ab}}$; adaxial traits are subscripted with $_{\textrm{ad}}$. In principle, the different relationship between traits on each surface could be caused by different evolutionary variance in stomatal density ($\text{Var}[\Delta \text{log}(D_\text{ab})] \ne \text{Var}[\Delta \text{log}(D_\text{ad})]$) and/or covariance ($\text{Cov}[\Delta \text{log}(L_\text{ab}), \Delta \text{log}(D_\text{ab})] \ne \text{Cov}[\Delta \text{log}(L_\text{ad}), \Delta \text{log}(D_\text{ad})]$) on each surface. However, the covariance relationship between density and length is similar on each surface, whereas the evolutionary variance in adaxial stomatal density is significantly higher than that for abaxial density ($\text{Var}[\Delta \text{log}(D_\text{ab})] < \text{Var}[\Delta \text{log}(D_\text{ad})]$; Fig. \ref{fig:h1}). Given that the average stomatal length is usually about the same on each surface (see below), these results imply that plants can often evolve stomatal densities on each surface without a concomitant change in size. Based on our theoretical analysis, we interpret these results to mean that cell divisions affecting stomatal index are less evolutionarily constrained than the asymmetric cell division preceding the guard cell meristemoid (Fig. \ref{fig:developmental-integration}; Table \ref{tab:predictions})

@@

<!-- # Author Contribution -->

<!-- CDM designed the study, compiled data, analyzed data, and wrote the manuscript with input from all authors. All authors contributed data.  -->

@@

The final data set and phylogeny used in the analysis are included in the Online Appendix. The raw anatomical data and source code are deposited on Dryad (https://datadryad.org/stash/share/Tk-7ZGwhdxyxAJu2dTsBdWDaId9qmog2fGYLWgGzPwU.)

@@

# Online Appendix

\renewcommand\thefigure{A\arabic{figure}}    
\renewcommand\thetable{A\arabic{table}}    
\renewcommand\theequation{A\arabic{equation}}    
\setcounter{figure}{0}    
\setcounter{table}{0}    
\setcounter{equation}{0}

>>>>

---
output:
  bookdown::pdf_document2:
    includes:
      before_body: ../template/amnat_doc_prefix.tex
      in_header: ../template/amnat_preamble.tex
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: no
    toc: no
  bookdown::html_document2:
    number_sections: no
    theme: readable
    toc: yes
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::word_document2: null
fontsize: 12pt
linestretch: 1.0
# linestretch: 1.7
link-citations: yes
bst: ../template/amnat.bst
bibliography: stomata-independence-1.bib
---

@@

<!-- % For Am Nat, moved to "template/amnat_preamble.txt" -->
\begin{flushleft}
{\Large
\textbf\newline{How important are functional and developmental constraints on phenotypic evolution? An empirical test with the stomatal anatomy of flowering plants}
}
\newline
\\
Christopher D. Muir\textsuperscript{1,*} (ORCID: 0000-0003-2555-3878),
Miquel \`{A}ngel Conesa\textsuperscript{2},
Jeroni Galm\'{e}s\textsuperscript{2} (ORCID: 0000-0002-7299-9349),
Varsha S. Pathare\textsuperscript{3} (ORCID: 0000-0001-6220-7531),
Patricia Rivera\textsuperscript{4},
Rosana López Rodríguez\textsuperscript{5},
Teresa Terrazas\textsuperscript{4},
Dongliang Xiong\textsuperscript{6} (ORCID: 0000-0002-6332-2627)
\\
\bigskip
\bf{1} School of Life Sciences, University of Hawaii at M\=anoa, Honolulu, HI 96822, USA \\
\bf{2} Research Group on Plant Biology under Mediterranean Conditions, Departament de Biologia, Universitat de les Illes Balears, Ctra. Valldemossa km 7.5, E-07122, Palma, Spain \\
\bf{3} School of Biological Sciences, Washington State University, Pullman, WA 99164-4236, USA \\
\bf{4} Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Apartado Postal 70‑367, 04510 Mexico City, Mexico \\
\bf{5} Departamento de Sistemas y Recursos Naturales, Universidad Politécnica de Madrid, 28040 Madrid, Spain \\
\bf{6} National Key Laboratory of Crop Genetic Improvement, MOA Key Laboratory of Crop Ecophysiology and Farming System in the Middle Reaches of the Yangtze River, College of Plant Science and Technology, Huazhong Agricultural University, Wuhan, Hubei 430070, China\\
\bigskip
$\ast$ cdmuir@hawaii.edu

\textit{Keywords}: adaptation, amphistomy, developmental integration, leaf, packing limits, phylogenetic comparative methods, stomata

@@

<!-- \textit{Manuscript elements}: Figure 1, figure 2, figure 3, figure 4, figure 5, figure 6, table 1, online supplement A (including Notes S1, notes S2, notes S3, figure S1, figure S2, figure S3, table S1, table S2, and table S3). -->

@@

<!-- \linenumbers{} -->
<!-- \modulolinenumbers[3] -->

\newpage{}

@@

Stomatal anatomy on the leaves of flowering plants provides an exceptional opportunity to test for functional and developmental constraints because 1) there are 1000s of species to compare and 2) the main packing constraints and developmental steps are analytically tractable. This means it is possible to derive quantitative predictions and test their generality using large comparative data sets representing millions of years of evolutionary history. For this purpose, a heretofore unappreciated fact about stomata is that many leaves have stomata on both lower and upper surfaces. The packing and developmental constraints are the same for stomata on each surface, but the selection may differ. Therefore, if packing and developmental constraints dominate, stomatal anatomy on each surface should diverge in concert. Failure to do so implies that independent evolution is possible and that selection against generally maladaptive phenotypes explains at least some of the covariance between traits. Phylogenetic comparisons of stomatal anatomy provide a statistically powerful, general, and elegant way to distinguish different phenotypic constraints that would be impossible in many other traits with as much ecological significance. The next sections provide background information on stomatal anatomy, how it varies, and why functional or developmental constraints might be important.

@@

@@

Given the significance of stomata for plant function and global vegetation modeling [@berry_stomata:_2010], we would like to understand what factors constrain their anatomical variation. Here we focus on interspecific variation in mean trait values rather than intraspecific variation. Packing constraints and developmental integration could explain inverse size-density scaling. The density and size of stomata on a planar leaf surface can be viewed as a packing problem where the total area allocated to stomata cannot exceed the total leaf area. This is a functional, or geometric, constraint because certain combinations of large size and high density are not physically possible. This functional constraint cannot explain why combinations of low density and small size are rare, but may explain why stomatal size must decrease when density increases as the leaf runs out of space. The packing limit of functional stomata is less than the entire leaf area, but the exact value is unclear. The realized upper limit is close to 1/3 or 1/2 [@de_boer_optimal_2016; @sack_developmental_2016; @liu_scaling_2021] for the species' mean, not an individual leaf.

@@

The second hypothesis is that cell size mediates developmental integration between stomatal size and density. If developmental integration is the primary reason for inverse size-density scaling, then amphistomatous leaves will exhibit identical size-density scaling on each surface. If the stomatal size and density scale differently on each surface, this implies that they can evolve independently and that selection against generally maladaptive trait combinations shapes their covariance. Furthermore, we predicted that divergence in genome size, which is strongly associated with meristematic cell volume [@simova_geometrical_2012], would covary with stomatal size and density similarly on each surface.

@@

@@

Unless otherwise mentioned, we performed all data wrangling and statistical analyses in *R* version `r as.character(getRversion())` [@r_core_team_r:_2022]. Source code is publicly available on GitHub (https://github.com/cdmuir/stomata-independence) and archived on Zenodo (see Data availability).

@@

Developmental integration could shape patterns of phenotypic macroevolution, but a major hindrance to progress is that verbal models do not make precise, quantitative predictions that distinguish it from alternatives. An advantage of testing developmental integration in stomata is that their development is well studied [@bergmann_stomatal_2007; @dow_patterning_2014; @sack_developmental_2016]. We can leverage that knowledge to build a developmental function and derive equations for phenotypic (co)variance caused by developmental integration. If observed patterns of evolution are inconsistent with developmental integration, theory may also help identify which parameters lead to developmental disintegration. We combined and extended stomatal development models to predict how stomatal density and length would diverge if stomatal development were constrained and how those predictions would change if stomatal development were unconstrained. We summarize our methods verbally here and direct readers interested in the mathematical details to Notes S1. A graphical summary is provided in Fig. \@ref(fig:developmental-integration). We imposed constraint by assuming the stomatal developmental function is constrained. The developmental function maps cell size prior to differentiation onto stomatal size and density using two parameters. The first parameter describes how cell volume is apportioned between epidermal cells and guard cell meristemoids during asymmetric cell division [@dow_patterning_2014]. The second parameter is stomatal index [@salisbury_i_1928; @sack_developmental_2016], which is determined by amplifying and spacing divisions after asymmetric cell division [@dow_patterning_2014]. When these parameters are fixed, divergence in stomatal size and density is determined by divergence in meristematic cell volume and expansion prior to asymmetric division. We relaxed this constraint by treating parameters of the developmental function as random variables that can diverge between species. We used random variable algebra to derive predicted (co)variance in divergence between stomatal length and density [see @lynch_genetics_1998 for key random variable algebra theorems].

@@

We searched the literature for studies that measured stomatal density and stomatal size, either guard cell length or stomatal pore length, for both abaxial and adaxial leaf surfaces. In other words, we did not include studies unless they reported separate density and size values for each surface. We did not record leaf angle because it is typically not reported, but we presume that for the vast majority of taxa that the abaxial is the lower surface and the adaxial is the upper surface. This is reversed in resupinate leaves, but to the best of our knowledge, our synthesis did not include resupinate leaves. None of the species with resupinate leaves listed by @chitwood_conflict_2012 are in our data set. We refer to guard cell length as stomatal length and converted stomatal pore length to stomatal length assuming guard cell length is twice pore length [@sack_developmental_2016]. Table \@ref(tab:traits) lists anatomical traits and symbols. Abaxial traits are subscripted with $_{\textrm{ab}}$; adaxial traits are subscripted with $_{\textrm{ad}}$

@@

We generated an ultrametric, bifurcating phylogeny of `r n_tax3` taxa by resolving and removing ambiguous taxonomic names, placing taxa on the GBOTB.extended mega-tree of seed plants [@smith_constructing_2018; @zanne_three_2014], and resolving polytomies using published sequence data. Divergence times in this phylogeny are based on extensive fossil calibration [see @smith_constructing_2018; @magallon_metacalibrated_2015 for further detail]. The complete methodology is described in the online supplement (Notes S2).

@@

<!-- Explanation for using pairs rather than whole tree. Removed based on feedback from Reviewer 2
We only used pairs of terminal taxa rather than the entire tree for two reasons. First, even in whole-tree methods, approximately half of the data comes from divergence at the tips, so we do not lose much statistical power. Second, using a taxon-pair method obviates making strong assumptions about multivariate trait evolution process homogeneity throughout flowering plant evolution. Homogeneity is biologically unlikely and it is difficult to fit more complex models reliably with only data from extant taxa.
-->

@@

All hypotheses make predictions about trait (co)variance matrices or parameters derived from them (see Notes S1 and subsections below). Within and among species covariation is a hallmark of developmental integration [@armbruster_multilevel_1988], but other evolutionary processes also lead to covariance. Distinguishing between them requires deriving predictions and testing whether observed covariance is consistent with one hypothesis or another. We estimated the $4 \times 4$ covariance matrix of phylogenetically independent contrasts between log-transformed values of $\Delta \text{log}(D_\mathrm{ab})$, $\Delta \text{log}(D_\mathrm{ad})$, $\Delta \text{log}(L_\mathrm{ab})$, and $\Delta \text{log}(L_\mathrm{ad})$ using a distributional multiresponse robust Bayesian approach. See Table \ref{tab:traits} for variable definitions. We denote variances as $\text{Var}[\Delta \text{log(trait)}]$ and covariances as $\text{Cov}[\Delta \text{log(trait}_1),\Delta \text{log(trait}_2)]$. We used a multivariate $t$-distribution rather than a Normal distribution because estimates using the former are more robust to exceptional trait values [@lange_robust_1989]. We also estimated whether the variance in trait divergence increases with time. Under many trait evolution models (e.g. Brownian motion), interspecific variance increases through time. To account for this, we included time since taxon-pair divergence as an explanatory variable affecting the trait covariance matrix.

@@

We asked how divergence in stomatal length and density would covary if the developmental function were constrained and compared it to their divergence when the developmental function can evolve. When the developmental function is constrained this means that allocation to guard cell meristemoids during asymmetric division and stomatal index are fixed (see Notes S1 for mathematical description). Under these assumptions, divergence in stomatal length and density is mediated entirely by divergence in meristematic cell volume and expansion prior to differentiation. Developmental integration is strong because divergence in density is perfectly negatively correlated with divergence in size. In contrast, stomatal length and density can diverge independently when the developmental function is not fixed. Divergence in asymmetric cell division affects stomatal size independently of density; divergence in stomatal index affects stomatal density independently of size. Divergence in the developmental function causes developmental disintegration because stomatal density and size can diverge independently. Developmental integration is minimal when asymmetric cell division and/or stomatal index diverge more than meristematic cell volume and expansion. The three main conclusions are that 1) developmental constraint leads to developmental integration; 2) different (co)variance in divergence of stomatal length and density on each surface implies the developmental function is not fixed; and 3) divergence in different components of the developmental function affect stomatal length and density differently. See Notes S1 and Table \@ref(tab:predictions) for more a complete derivation and detailed predictions.

@@

@@

@@

@@

Functional and developmental constraints may slow adaptation by preventing traits from evolving independently towards a multivariate phenotypic optimum. An alternative view is that phenotypic evolution is limited primarily because most phenotypes are maladaptive and selected against, sometimes referred to as selective constraint. In this study, we took advantage of the fact that amphistomatous leaves produce stomata on both abaxial (usually lower) and adaxial (usually upper) surfaces. Packing limits, a functional constraint, and developmental integration should result in similar (co)variance in divergence of stomatal traits on each surface (Notes S1), whereas differing selective pressures for each surface would lead to different patterns of divergence. Although some patterns of trait (co)variance are compatible with packing limits or developmental integration (see below), independent evolution of stomatal density on each surface across flowering plants is notably inconsistent with these constraints. We therefore conclude that selection against generally maladaptive trait combinations plays a major role in the evolution of stomata and possibly other ecologically important traits. It should not be surprising that stomatal traits respond to natural selection, but it is nevertheless plausible that some stomatal trait combinations that would have high fitness are precluded by nonadaptive constraints. Our results challenge these nonadaptive explanations, especially for stomatal density.

@@

If meristematic cell volume and expansion integrate stomatal size and density [@brodribb_unified_2013], then we predicted inverse size-density scaling would evolve with the same (co)variance for both ab- and adaxial leaf surfaces (Notes S1). Contrary to this prediction, there are many combinations of stomatal density ($D_i$) and length ($L_i$) found on adaxial leaf surfaces that are absent from abaxial leaf surfaces (Fig. \ref{fig:h1_raw}). In principle, the different relationship between traits on each surface could be caused by different evolutionary variance in stomatal density ($\text{Var}[\Delta \text{log}(D_\text{ab})] \ne \text{Var}[\Delta \text{log}(D_\text{ad})]$) and/or covariance ($\text{Cov}[\Delta \text{log}(L_\text{ab}), \Delta \text{log}(D_\text{ab})] \ne \text{Cov}[\Delta \text{log}(L_\text{ad}), \Delta \text{log}(D_\text{ad})]$) on each surface. However, the covariance relationship between density and length is similar on each surface, whereas the evolutionary variance in adaxial stomatal density is significantly higher than that for abaxial density ($\text{Var}[\Delta \text{log}(D_\text{ab})] < \text{Var}[\Delta \text{log}(D_\text{ad})]$; Fig. \ref{fig:h1}). Given that the average stomatal length is usually about the same on each surface (see below), these results imply that plants can often evolve stomatal densities on each surface without a concomitant change in size. Based on our theoretical analysis, we interpret these results to mean that cell divisions affecting stomatal index are less evolutionarily constrained than the asymmetric cell division preceding the guard cell meristemoid (Fig. \ref{fig:developmental-integration}; Table \ref{tab:predictions})

@@

# Statement of Authorship

CDM led conceptualization, data analysis, and writing the original draft. All authors collected data and assisted with review and editing.

@@

The final data set and phylogeny used in the analysis are included in the Online Supplement. The raw anatomical data and source code are deposited in the Dryad Digital Repository: https://doi.org/10.5061/dryad.0rxwdbs42 [@muir_data_2022].

@@

# Tables
